NHL Realignment Effects on Playoff Qualification

### 2022-02-15 Paper.Rmd
Fixed spelling errors

# Motivation
   In the 2013-14 National Hockey League (NHL) season, the league made some changes to the conference numbers. The Winnipeg Jets moved from the Eastern conference to the Western conference. This created an imbalance in the conference numbers giving the Eastern conference sixteen teams and leaving the Western conference with only fourteen teams. From the start, it already seemed unfair to have more teams in one conference than the other. If every team had equal skill, the fourteen teams in the western conference had an equal one four in seven chance of making it to the playoffs. In contrast, the sixteen teams in the eastern conference had a fifty percent chance of making it to the playoffs. This uneven probabillity will be known for the rest of the paper as the conference gap.

  Now how could an organization as big as the NHL allow for that kind of discrepency between the conferences? As a continuation of our research we wanted to look at another season with a similiar balance shift in the conferences to see how that change could have affected the outcome of the season. The season we chose was the recent expansion draft season of 2017-18. Before the start of the season, the NHL added a new franchise in the Las Vegas Golden Knights. This team was set to join the Western conference. This shifted the conference split from fourteen to sixtten West to East to a closer fifteen to sixteen split. We wanted to see if our result was statistically significant from both the conference gap of an even season and the conference gap from our tested 2013-14 season.

## Introduction
  In 2011 the Atlanta Thrashers became the Winnipeg Jets. A problem was presented as Winnipeg is considerably far from the east coast of the North American continent. Because of this and how the season is set up, the average travel time for teams playing winnipeg were fairly large as eastern conference teams play each other more often. Just before the start of the 2013-14 season, the league offices for the NHL decided to move over the Winnipeg Jets from the Eastern conference to the Western conference and in return the Eastern conference aquired the Columbus Blue Jackets and the Detroit Redwings. This left the the two conferences at an uneven balance. The Eastern conference now had sixteen teams while the Western conference only had fourteen teams. This presented a curious problem as for both conferences, eight teams make the playoff brackets. It is easy to see the problem at hand that the western conference has just as many playoff slots for having two fewer teams. The goal of our analysis and the paper we based our ideas off of and partially replicated, was to discover what that difference could mean in the greater sense of the league moving forward with this set up. We did this based on the average points and wins necessary to make the playoffs for both conferences.

  To further our study, we also looked at the 2017-18 season. This season is important because it gave the introduction of a new team in the league known as the Vegas Golden Knights. This team was introduced into the Pacific division of the Western conference, this meant that the balance between the number of teams in the East and West was more equal than the 2013-2014 split. Our goal for this was to investigate how this new conference structure could affect the conference gap. We used the same process to investigate this new season's structure as our previous analysis of the 2013-14 season. Given that the difference in the number of teams between conferences is less severe we suspected that our estimated conference gap under this setup (3 Divisions of 8 teams, 1 Division of 7 teams) would lie somewhere around the halfway point between the estimate for the pre 2013-2014 structure (6 Divisions of 5 teams each) and the 2013-2014 season (2 Divisions of 8 teams, 2 Divisions of 7 teams).

>>>>

# Motivation
  Before the 2013-14 National Hockey League (NHL) season, the league made changes to the number of teams in each of the two conferences. The Winnipeg Jets moved from the Eastern Conference to the Western Conference which created an imbalance in the conference numbers giving the Eastern Conference sixteen teams and leaving the Western Conference with only fourteen teams. Since eight teams from each conference qualify for the playoffs, this imbalance created a “conference gap” that affected the probability of a team being selected for the playoffs.

  How much would this change affect the outcome of the playoff selection process? This question was researched by Pettigrew1 in 2014 using a Monte Carlo approach and found the conference gap to be on average 2.74 points, meaning that teams in the Eastern Conference will have to win 1-2 games more than teams in the Western Conference in order to make the playoffs. For our research we validate Pettigrew’s paper as well as extend it by using a similar approach to compare this 2013-2014 season to another season with a similar balance shift in the conferences to see how the changes compare statistically.

  The season we chose was the recent expansion draft season of 2017-18. Before the start of the season, the NHL added a new franchise in the Las Vegas Golden Knights which was set to join the Western conference. This added a team to the Western conference and resulted in a team split of fifteen to sixteen between East and West.  Our goal was to see if the change in a team’s probability of making the playoffs was statistically significant when comparing between a year with even team distribution and the conference gap year from our tested 2013-14 and 2017-18  seasons.

## Introduction
  In 2011 the Atlanta Thrashers of the Eastern Conference became the Winnipeg Jets. This presented a problem as Winnipeg is a considerable distance from the east coast. Because of this and how the season game schedule is set up, the average travel time for teams playing Winnipeg from the Eastern Conference would have been fairly large. So, just before the start of the 2013-14 season, the league offices for the NHL decided to move the Winnipeg Jets from the Eastern Conference to the Western Conference and in return the Eastern Conference acquired the Columbus Blue Jackets and the Detroit Redwings. This left the two conferences with a different number of teams. The Eastern Conference then had sixteen teams while the Western Conference only had fourteen teams. This presented a curious problem since both conferences had eight teams advance to the playoffs at the end of the season.

  The goal of our analysis and the paper we based our ideas off of and partially replicated1, was to discover how this impacted a team’s chances of making the playoffs. We did this based on the average points and wins necessary to make the playoffs for both conferences.

  To further our study, we also looked at the 2017-18 season. This season is important because it gave the introduction of a new team in the league known as the Vegas Golden Knights. This team was introduced into the Pacific division of the Western conference, which meant that the balance between the number of teams in the East and West was more equal than the 2013-2014 split. Our goal for this was to investigate how this new conference structure could affect the conference gap. We used the same process to investigate this new season’s structure as our previous analysis of the 2013-14 season. Given that the difference in the number of teams between conferences is less severe we suspected that our estimated conference gap under this setup (3 Divisions of 8 teams, 1 Division of 7 teams) would lie somewhere around the halfway point between the estimate for the pre 2013-2014 structure (6 Divisions of 5 teams each) and the 2013-2014 season (2 Divisions of 8 teams, 2 Divisions of 7 teams).

### 2022-02-15 Paper.tex
Fixed spelling errors

---
title: "NHL Realignment Effects on Playoff Qualification"
author: "Kilbourne Charrier, Michael Bauers, Clinton Rothschild, Sam Downs"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Motivation
   In the 2013-14 National Hockey League (NHL) season, the league made some changes to the conference numbers. The Winnipeg Jets moved from the Eastern conference to the Western conference. This created an imbalance in the conference numbers giving the Eastern conference sixteen teams and leaving the Western conference with only fourteen teams. From the start, it already seemed unfair to have more teams in one conference than the other. If every team had equal skill, the fourteen teams in the western conference had an equal one four in seven chance of making it to the playoffs. In contrast, the sixteen teams in the eastern conference had a fifty percent chance of making it to the playoffs. This uneven probabillity will be known for the rest of the paper as the conference gap.

@@

## Introduction
  In 2011 when the Atlanta Thrashers became the Winnipeg Jets. A problem was presented as Winnipeg is considerably far from the east coast of the North American continent. Because of this and how the season is set up, the average travel time for teams playing winnipeg were fairly large as eastern conference teams play each other more often. Just before the start of the 2013-14 season, the league offices for the NHL decided to move over the Winnipeg Jets from the Eastern conference to the Western conference and in return the Eastern conference aquired the Columbus Blue Jackets and the Detroit Redwings. This left the the two conferences at an uneven balance. The Eastern conference now had sixteen teams while the Western conference only had fourteen teams. This presented a curious problem as for both conferences, eight teams make the playoff brackets. It is easy to see the problem at hand that the western conference has just as many playoff slots for having two less teams. The goal of our analysis and the paper we based our ideas off of and partially replicated, was to discover what that difference could mean in the grander sense of the league moving forward with this set up. We did this based on the average points and wins necessary to make the playoffs for both conferences.

@@

## Pre 2013-2014 NHL Conference Structure

![](Images/pre2013struct.png){width=60%}

## 2013-2014 NHL Conference Structure

![](Images/2013struct.png){width=60%}

## 2017-2018 NHL Conference Structure

![](Images/2017struct.png){width=60%}

# Methodology

## Basic Idea

The basic idea behind the model is that all of the teams have a base "skill level" that they play at or near during any given game throughout the season. Each game is competitive because we aren't sure how well either team will play, and even a team with a higher skill level can be beaten by another team for any number of reasons. We keep track of this skill level and draw other metrics based off of it, then use the metrics to compute a team's score for the season. This score is used to calculate the team's seed in their respective conference and division.

## Creating the Model

To start out, we first need to make sure that we calculate the game-to-game performance levels given by $\tau$ and the standard deviation of the team's talent levels given by $\sigma$. These are related to the following ratio, where p is the probability that the worst team in the league beats the best team in the league.

$-\frac{2.8854}{\Phi^{-1}(p)} = \frac{\tau}{\sigma}$

Based on the data from all of the seasons with 30 teams, we can calculate that the team with the worst record in the league beat the team with the best team about a quarter of the time. We choose to set p equal to 0.25 for this reason, making the ratio $\tau/\sigma$ = 4.28. This means that there are four times the variability in the team's performances in game than there is in the talent levels of the teams. The actual values of $\tau$ and $\sigma$ don't matter, just the ratio, but for simplicity we will set $\tau$ equal to 4.28 and $\sigma$ equal to 1.

We then want to simulate a skill level metric for each of the thirty teams. We create the metric from the following distribution, with team i being equal to $\mu_{i}$.

$\mu_{i}$~$N(0,\sigma^{2})$

@@

For each game that is played in the season we draw a performance metric for both competing teams. This metric is drawn from a normal distribution with the mean being equal to the team's skill level and the standard deviation being equal to $\tau^{2}$. This is based on the idea that a team with any skill level will play somewhere around their skill level, rarely playing much worse and rarely playing much better. This metric is calculated as such, with i indicating the home team and j indicating the away team.

$\gamma_{i}$~$N(\mu_{i}, \tau^{2})$

$\gamma_{j}$~$N(\mu_{j}, \tau^{2})$

Logically, the winner of each game would be decided by the higher value of $\gamma$, but in hockey it is possible for the game to go into overtime. This happens in about 22.4% of the games that were played over the past 15 seasons, so we must solve the following equasion for $\alpha$:

$Pr(|\gamma_{i}-\gamma_{j}| < \alpha) = 22.4$%

Solving this equasion gives us a value of $\alpha \approx 1.769.$

If there is a game in which $|\gamma_{i}-\gamma_{j}|\geq \alpha$, the team with the higher $\gamma$ value is awarded a win and is therefore given two points while the team with the lower $\gamma$ value is given 0 points. For games in which $|\gamma_{i}-\gamma_{j}| < \alpha$, the game is considered a tie.

In the case of a tie we linearly resale the $\gamma$ values with the following formula:

$\zeta = \frac{(\gamma_{i}-\gamma_{j}) - a}{b}$

Where a is the mean and b is the maximum of all $\gamma_{i}-\gamma_{j}$ values from that season.

Using the resulting value of $\zeta$ we use a Bernoulli distribution to simulate a weighted coin flip, with a success meaning that the home team wins and a failure meaning the away team wins.

Winner ~ $Bernoulli(\zeta_{ij})$

@@

## Additional model details

Our model does not account for the real-world skill of teams, but because we are not doing any inference on specific team performance depending on rule changes this is not relevant. Because we are interested in the difference in scores of the 8th seeds we actually don't even need any real-world teams and can instead replace them with placeholders or just numbers. The lack of a need for real world data was also an advantage to our analysis as it meant our results may be expressed as a general representation of the "fairness" of these conference structures since they are not related to the skill of real world teams.

An alternative choice we could have made is that in the event of a tie we could have given either team a 50% chance of winning instead of linearly rescaling the values and using a weighted coin flip, but we don't know exactly how the outcome of NHL overtime games usually end up skewing -- whether it be in favor of the better team or if it is distributed more randomly.

When performing analysis on the 2017-2018 conference structure we chose to hold our parameters $p,\tau$ and $\sigma$ to be the same as stated above in the 2013-2014 analysis since we are talking about a reasonable small amount of time passing between these two seasons we thought it would be unreasonable to think that they would alter all that much. The same number of simulations were run also to make results more easily comparable.

# Results

@@

![](Images/GapVis.png){width=85%}

@@

![](Images/gapvis20172018.png){width=85%}

@@

The mean conference gap, in terms of total points, means that the East’s 8th seeded team has on average 2.30 more points than the West’s 8th seeded team for 2013-2014, similarly we expect the 8th seed in the Eastern conference to have 1.091 more points than their Western conference counterpart under the 2017-2018 structure. In the NHL, a team earns 2 points for a win, 1 point for an overtime loss, and 0 points for a loss. With this point system,under the 2013-2014 alignment the East’s 8th seeded team on average needs to win one more game during the season or exchange two losses for two losses in overtime. This favors Western conference teams in the way they have to earn less total points throughout the season to make the playoffs than Eastern conference teams. This same bias towards Western conference teams to make the playoffs is present under the 2017-2018 bias but the practical effects of a one point difference is not even worth a single win in an 82 game season, so although this effect is present it's not as drastic.

Not only were the 8th seeds from each conference calculated but each seed for the whole NHL. These box and whisker plots show a wide range of information for the total points for each seed across the 10,000 simulated seasons for both the 2013-2014 and 2017-2018 alignemnts. The endpoints of the whiskers represent the outlier point totals earned by each seed. The sides of the box represent the 2.5% and 97.5% quantiles. The line within the box represents the mean. The structure of this plot makes sense with the curve of less points the lower seed and more points the higher seed the team achieves. The higher the number on the x axis the lower seed of the team; 30 being the worst team in the league (31 for 2017-2018 simulation) and 1 being the best team in the league. Interesting feautures of these plots are a tight spread of simulated points for teams seeded around the middle of the league. This makes sense given that these point values are probably closer to the mean point total for a season and thus more likely under the model we defined. Another note is that the overall range of point totals is greater in the 2013-2014 simulation than it is under the 2017-2018 simulation. This has been compared with real NHL data and the shape lines up in a similar fashion.

## 2013-2014 Points by Seed
![](Images/SeedPlace.png){width=80%}

## 2017-2018 Points by Seed
![](Images/seedplace20172018.png){width=80%}

# Conclusion

Since the change to the conference structure of the NHL to be unbalanced, there have been questions on whether these structures are giving an unfair advantage for one of the conferences: more specifically the Western Conference. With the West having two less teams under the 2013-2014 alignment, it was inferred that each of those teams would have a better chance of making the playoffs. They wouldn’t need to earn as many points to secure a playoff spot as teams in the Eastern Conference. After running 10,000 Monte Carlo simulations of the 2013-2014 NHL season, the results support the case that the Western Conference has a noticeable advantage over the Eastern Conference. The 8th seed in the east on average has an end of season point total of 2.3 (1.091 for 2017-2018) points higher than the 8th seed in the west. This can cause problems because a statistically worse team can make the playoffs in the Western Conference while a better team might miss out in the Eastern Conference. This is support of the idea that these rules introduce a bias agaisnt teams in the Eastern conference which goes agaisnt the league being "fair" to all teams.

# References

- Pettigrew, Stephen. “How the West Will Be Won: Using Monte Carlo Simulations to Estimate the Effects of NHL Realignment.” Journal of Quantitative Analysis in Sports, vol. 10, no. 3, 1 Jan. 2014, 10.1515/jqas-2013-0125. Accessed 10 Dec. 2020.

- History of Organizational Changes in the NHL. 26 Nov. 2020, en.wikipedia.org/wiki/History_of_organizational_changes_in_the_NHL. Accessed 10 Dec. 2020.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\title{NHL Realignment Effects on Playoff Qualification}
\author{Kilbourne Charrier, Michael Bauers, Clinton Rothschild, Sam
Downs}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={NHL Realignment Effects on Playoff Qualification},
  pdfauthor={Kilbourne Charrier, Michael Bauers, Clinton Rothschild, Sam Downs},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\begin{document}
\maketitle

\hypertarget{motivation}{%
\section{Motivation}\label{motivation}}

In the 2013-14 National Hockey League (NHL) season, the league made some
changes to the conference numbers. The Winnipeg Jets moved from the
Eastern conference to the Western conference. This created an imbalance
in the conference numbers giving the Eastern conference sixteen teams
and leaving the Western conference with only fourteen teams. From the
start, it already seemed unfair to have more teams in one conference
than the other. If every team had equal skill, the fourteen teams in the
western conference had an equal one four in seven chance of making it to
the playoffs. In contrast, the sixteen teams in the eastern conference
had a fifty percent chance of making it to the playoffs. This uneven
probabillity will be known for the rest of the paper as the conference
gap.

@@

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

In 2011 the Atlanta Thrashers became the Winnipeg Jets. A problem was
presented as Winnipeg is considerably far from the east coast of the
North American continent. Because of this and how the season is set up,
the average travel time for teams playing winnipeg were fairly large as
eastern conference teams play each other more often. Just before the
start of the 2013-14 season, the league offices for the NHL decided to
move over the Winnipeg Jets from the Eastern conference to the Western
conference and in return the Eastern conference aquired the Columbus
Blue Jackets and the Detroit Redwings. This left the the two conferences
at an uneven balance. The Eastern conference now had sixteen teams while
the Western conference only had fourteen teams. This presented a curious
problem as for both conferences, eight teams make the playoff brackets.
It is easy to see the problem at hand that the western conference has
just as many playoff slots for having two fewer teams. The goal of our
analysis and the paper we based our ideas off of and partially
replicated, was to discover what that difference could mean in the
greater sense of the league moving forward with this set up. We did this
based on the average points and wins necessary to make the playoffs for
both conferences.

@@

\hypertarget{pre-2013-2014-nhl-conference-structure}{%
\subsection{Pre 2013-2014 NHL Conference
Structure}\label{pre-2013-2014-nhl-conference-structure}}

\includegraphics[width=0.6\textwidth,height=\textheight]{Images/pre2013struct.png}

\hypertarget{nhl-conference-structure}{%
\subsection{2013-2014 NHL Conference
Structure}\label{nhl-conference-structure}}

\includegraphics[width=0.6\textwidth,height=\textheight]{Images/2013struct.png}

\hypertarget{nhl-conference-structure-1}{%
\subsection{2017-2018 NHL Conference
Structure}\label{nhl-conference-structure-1}}

\includegraphics[width=0.6\textwidth,height=\textheight]{Images/2017struct.png}

\hypertarget{methodology}{%
\section{Methodology}\label{methodology}}

\hypertarget{basic-idea}{%
\subsection{Basic Idea}\label{basic-idea}}

The basic idea behind the model is that all of the teams have a base
``skill level'' that they play at or near during any given game
throughout the season. Each game is competitive because we aren't sure
how well either team will play, and even a team with a higher skill
level can be beaten by another team for any number of reasons. We keep
track of this skill level and draw other metrics based off of it, then
use the metrics to compute a team's score for the season. This score is
used to calculate the team's seed in their respective conference and
division.

\hypertarget{creating-the-model}{%
\subsection{Creating the Model}\label{creating-the-model}}

To start out, we first need to make sure that we calculate the
game-to-game performance levels given by \(\tau\) and the standard
deviation of the team's talent levels given by \(\sigma\). These are
related to the following ratio, where p is the probability that the
worst team in the league beats the best team in the league.

\(-\frac{2.8854}{\Phi^{-1}(p)} = \frac{\tau}{\sigma}\)

Based on the data from all of the seasons with 30 teams, we can
calculate that the team with the worst record in the league beat the
team with the best team about a quarter of the time. We choose to set p
equal to 0.25 for this reason, making the ratio \(\tau/\sigma\) = 4.28.
This means that there are four times the variability in the team's
performances in game than there is in the talent levels of the teams.
The actual values of \(\tau\) and \(\sigma\) don't matter, just the
ratio, but for simplicity we will set \(\tau\) equal to 4.28 and
\(\sigma\) equal to 1.

We then want to simulate a skill level metric for each of the thirty
teams. We create the metric from the following distribution, with team i
being equal to \(\mu_{i}\).

\(\mu_{i}\)\textasciitilde{}\(N(0,\sigma^{2})\)

@@

For each game that is played in the season we draw a performance metric
for both competing teams. This metric is drawn from a normal
distribution with the mean being equal to the team's skill level and the
standard deviation being equal to \(\tau^{2}\). This is based on the
idea that a team with any skill level will play somewhere around their
skill level, rarely playing much worse and rarely playing much better.
This metric is calculated as such, with i indicating the home team and j
indicating the away team.

\(\gamma_{i}\)\textasciitilde{}\(N(\mu_{i}, \tau^{2})\)

\(\gamma_{j}\)\textasciitilde{}\(N(\mu_{j}, \tau^{2})\)

Logically, the winner of each game would be decided by the higher value
of \(\gamma\), but in hockey it is possible for the game to go into
overtime. This happens in about 22.4\% of the games that were played
over the past 15 seasons, so we must solve the following equasion for
\(\alpha\):

\(Pr(|\gamma_{i}-\gamma_{j}| < \alpha) = 22.4\)\%

Solving this equasion gives us a value of \(\alpha \approx 1.769.\)

If there is a game in which \(|\gamma_{i}-\gamma_{j}|\geq \alpha\), the
team with the higher \(\gamma\) value is awarded a win and is therefore
given two points while the team with the lower \(\gamma\) value is given
0 points. For games in which \(|\gamma_{i}-\gamma_{j}| < \alpha\), the
game is considered a tie.

In the case of a tie we linearly resale the \(\gamma\) values with the
following formula:

\(\zeta = \frac{(\gamma_{i}-\gamma_{j}) - a}{b}\)

Where a is the mean and b is the maximum of all
\(\gamma_{i}-\gamma_{j}\) values from that season.

Using the resulting value of \(\zeta\) we use a Bernoulli distribution
to simulate a weighted coin flip, with a success meaning that the home
team wins and a failure meaning the away team wins.

Winner \textasciitilde{} \(Bernoulli(\zeta_{ij})\)

@@

\hypertarget{additional-model-details}{%
\subsection{Additional model details}\label{additional-model-details}}

Our model does not account for the real-world skill of teams, but
because we are not doing any inference on specific team performance
depending on rule changes this is not relevant. Because we are
interested in the difference in scores of the 8th seeds we actually
don't even need any real-world teams and can instead replace them with
placeholders or just numbers. The lack of a need for real world data was
also an advantage to our analysis as it meant our results may be
expressed as a general representation of the ``fairness'' of these
conference structures since they are not related to the skill of real
world teams.

An alternative choice we could have made is that in the event of a tie
we could have given either team a 50\% chance of winning instead of
linearly rescaling the values and using a weighted coin flip, but we
don't know exactly how the outcome of NHL overtime games usually end up
skewing -- whether it be in favor of the better team or if it is
distributed more randomly.

When performing analysis on the 2017-2018 conference structure we chose
to hold our parameters \(p,\tau\) and \(\sigma\) to be the same as
stated above in the 2013-2014 analysis since we are talking about a
reasonable small amount of time passing between these two seasons we
thought it would be unreasonable to think that they would alter all that
much. The same number of simulations were run also to make results more
easily comparable.

\hypertarget{results}{%
\section{Results}\label{results}}

@@

\includegraphics[width=0.85\textwidth,height=\textheight]{Images/GapVis.png}

@@

\includegraphics[width=0.85\textwidth,height=\textheight]{Images/gapvis20172018.png}

@@

The mean conference gap, in terms of total points, means that the East's
8th seeded team has on average 2.30 more points than the West's 8th
seeded team for 2013-2014, similarly we expect the 8th seed in the
Eastern conference to have 1.091 more points than their Western
conference counterpart under the 2017-2018 structure. In the NHL, a team
earns 2 points for a win, 1 point for an overtime loss, and 0 points for
a loss. With this point system,under the 2013-2014 alignment the East's
8th seeded team on average needs to win one more game during the season
or exchange two losses for two losses in overtime. This favors Western
conference teams in the way they have to earn less total points
throughout the season to make the playoffs than Eastern conference
teams. This same bias towards Western conference teams to make the
playoffs is present under the 2017-2018 bias but the practical effects
of a one point difference is not even worth a single win in an 82 game
season, so although this effect is present it's not as drastic.

Not only were the 8th seeds from each conference calculated but each
seed for the whole NHL. These box and whisker plots show a wide range of
information for the total points for each seed across the 10,000
simulated seasons for both the 2013-2014 and 2017-2018 alignemnts. The
endpoints of the whiskers represent the outlier point totals earned by
each seed. The sides of the box represent the 2.5\% and 97.5\%
quantiles. The line within the box represents the mean. The structure of
this plot makes sense with the curve of less points the lower seed and
more points the higher seed the team achieves. The higher the number on
the x axis the lower seed of the team; 30 being the worst team in the
league (31 for 2017-2018 simulation) and 1 being the best team in the
league. Interesting feautures of these plots are a tight spread of
simulated points for teams seeded around the middle of the league. This
makes sense given that these point values are probably closer to the
mean point total for a season and thus more likely under the model we
defined. Another note is that the overall range of point totals is
greater in the 2013-2014 simulation than it is under the 2017-2018
simulation. This has been compared with real NHL data and the shape
lines up in a similar fashion.

\hypertarget{points-by-seed}{%
\subsection{2013-2014 Points by Seed}\label{points-by-seed}}

\includegraphics[width=0.8\textwidth,height=\textheight]{Images/SeedPlace.png}

\hypertarget{points-by-seed-1}{%
\subsection{2017-2018 Points by Seed}\label{points-by-seed-1}}

\includegraphics[width=0.8\textwidth,height=\textheight]{Images/seedplace20172018.png}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

Since the change to the conference structure of the NHL to be
unbalanced, there have been questions on whether these structures are
giving an unfair advantage for one of the conferences: more specifically
the Western Conference. With the West having two less teams under the
2013-2014 alignment, it was inferred that each of those teams would have
a better chance of making the playoffs. They wouldn't need to earn as
many points to secure a playoff spot as teams in the Eastern Conference.
After running 10,000 Monte Carlo simulations of the 2013-2014 NHL
season, the results support the case that the Western Conference has a
noticeable advantage over the Eastern Conference. The 8th seed in the
east on average has an end of season point total of 2.3 (1.091 for
2017-2018) points higher than the 8th seed in the west. This can cause
problems because a statistically worse team can make the playoffs in the
Western Conference while a better team might miss out in the Eastern
Conference. This is support of the idea that these rules introduce a
bias agaisnt teams in the Eastern conference which goes agaisnt the
league being ``fair'' to all teams.

\hypertarget{references}{%
\section{References}\label{references}}

\begin{itemize}
\item
  Pettigrew, Stephen. ``How the West Will Be Won: Using Monte Carlo
  Simulations to Estimate the Effects of NHL Realignment.'' Journal of
  Quantitative Analysis in Sports, vol.~10, no. 3, 1 Jan.~2014,
  10.1515/jqas-2013-0125. Accessed 10 Dec.~2020.
\item
  History of Organizational Changes in the NHL. 26 Nov.~2020,
  en.wikipedia.org/wiki/History\_of\_organizational\_changes\_in\_the\_NHL.
  Accessed 10 Dec.~2020.
\end{itemize}

\end{document}
